\textbf{ Spectrophotometric parallax:\\ Accurate estimates of luminosity and distance\\ for luminous red-giant stars }

### 2018-08-01 paper/sp.tex
notes2self
thanks ACE

% To-do items
% -----------
% - draft introduction
% - draft discussion
% - make sure we address the sensibleness and generality of the linear model for the photometric space
% - address the non-convexity of the model (see twitter convo with @andy_l_jones).
% - Coryn's point that we should carefully explain how our model is geometric.
% - related: HWR's point that this is standardizeable candles, like the Ia SNe results.
% - related: Put in an explanation of WHY this works so well.
% - related: Put in an explanation of how it is that we aren't really modeling luminosity or distance but really just generating parallaxes.
% - related: Switch the model to be a model for parallax directly.
% - does the model need a cute name, like The Swan?

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\sigma_n^2}
\quad ,
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\sigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 8000-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 8000.

@@

Because the \apogee\ target selection is based on \zmass, every \apogee\ star
also comes with \zmass photometry.
That is, for each star $n$,
we have near-infrared photometry $J_n$, $H_n$, and $K_n$.
In detail we use from the catalog the entries ACE WHAT SPECIFIC CATALOG ENTRIES.

@@

We take from the \gaia\ \acronym{DR2} Catalog the photometric data $G_n$
(\code{phot_g_mean_mag}) and $({BP}-{RP})_n$ (\code{bp_rp}), which will
become part of the feature vector $x_n$.
We do not use the entire \gaia\ \acronym{DR2} Catalog but instead the subset
that meets ACE WHAT ADDITIONAL PHOTOMETRIC CUTS?

We need complete feature-vector information for all stars.  For this
reason, we define the Parent Sample to be the set of all stars that
meet the \apogee\ and \gaia\ cuts and also
have the complete set of photometry: $G_n$, $({BP}-{RP})_n$, $J_n$,
$H_n$, $K_n$, $W_{1n}$, and $W_{2n}$ and also an apogee spectrum.
This Parent Sample contains ACE HOW MANY STARS? and is shown in HOGG WHAT FIGURES?

@@

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, ({BP}-{RP})_n, J_n, H_n, K_n, W_{1n}, W_{2n}, f_{1n}, f_{2n}, \cdots, f_{Ln}]
\quad ,
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
and
the fluxes in the $L=$ACE \apogee\ spectral pixels for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on.
These feature vectors live in a $D$-dimensional space where $D=$HOGG.
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured parallax $\gparallax_n$
and uncertainty $\sigma_n$.

@@

\acknowledgements
It is a pleasure to thank
  Coryn Bailer-Jones (\acronym{MPIA}),
  Andy L. Jones,
  Daniel Michalik (\acronym{ESTEC}), and
  Melissa Ness (Columbia)
for help with this project.
This project was developed in part at the
2018 \acronym{NYC} Gaia Sprint, hosted by the Center for Computational Astrophysics of
the Flatiron Institute in New York City in 2018 June.

>>>>

% To-do items
% -----------
% - draft introduction
% - draft discussion
% - make sure we address the sensibleness and generality of the linear model for the photometric space
% - address the non-convexity of the model (see twitter convo with @andy_l_jones).
% - Coryn's point that we should carefully explain how our model is geometric.
% - related: HWR's point that this is standardizeable candles, like the Ia SNe results.
% - related: Put in an explanation of WHY this works so well.
% - related: Put in an explanation of how it is that we aren't really modeling luminosity or distance but really just generating parallaxes.
% - does the model need a cute name, like The Swan?
% - we need a Facilities list / tags
% - we need a Software list / tags

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\sigma_n^2}
\quad ,
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\sigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 7400-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 7400.

@@

Because the \apogee\ target selection is based on \zmass, every \apogee\ star
also comes with \zmass\ (HOGG CITE) photometry.
That is, for each star $n$,
we have \zmass\ near-infrared photometry $J_n$, $H_n$, and $K_n$.

@@

We take from the \gaia\ \acronym{DR2} Catalog the photometric data
$G_n$ (\code{phot_g_mean_mag}),
${BP}_n$ (\code{phot_bp_mean_mag}),
and ${RP})_n$ (\code{phot_rp_mean_mag}),
which will become part of the feature vector $x_n$.
We do not use the entire \gaia\ \acronym{DR2} Catalog but instead the subset
that meets ACE WHAT ADDITIONAL PHOTOMETRIC CUTS?

We need complete feature-vector information for all stars.  For this
reason, we define the Parent Sample to be the set of all stars that
meet the \apogee\ and \gaia\ cuts and also
have the complete set of photometry: $G_n$, ${BP}_n$, ${RP}_n$, $J_n$,
$H_n$, $K_n$, $W_{1n}$, and $W_{2n}$ and also an apogee spectrum.
This Parent Sample contains ACE HOW MANY STARS? and is shown in HOGG WHAT FIGURES?

@@

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, {BP}_n, {RP}_n, J_n, H_n, K_n, W_{1n}, W_{2n}, f_{1n}, f_{2n}, \cdots, f_{Ln}]
\quad ,
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
and
the fluxes in the $L=7405$ \apogee\ spectral pixels (for which there are reliably
and consistently data) for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on.
These feature vectors live in a $D$-dimensional space where $D=7414$
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured parallax $\gparallax_n$
and uncertainty $\sigma_n$.

@@

\acknowledgements
It is a pleasure to thank
  Coryn Bailer-Jones (\acronym{MPIA}),
  Andy L. Jones,
  Daniel Michalik (\acronym{ESTEC}),
  Melissa Ness (Columbia),
  and the participants in the \acronym{MPIA} Milky Way Group Meeting
for help with this project.
This project was developed in part at the
2018 \acronym{NYC} Gaia Sprint, hosted by the Center for Computational Astrophysics of
the Flatiron Institute in New York City in 2018 June.
